Arrows for Parallel Computations

### 2017-09-26 paper/content/benchmarks.tex
minor
benchmarks missing for now.
replaced "Multicore Haskell" with "GpH"

\newcommand{\rmtest}{Rabin--Miller test\xspace}
\newcommand{\sudokutest}{Sudoku\xspace}
\newcommand{\jacobitest}{Jacobi sum test\xsp}
\newcommand{\torustest}{Gentleman\xspace}
\newlength{\plotwidthSMP}
\setlength{\plotwidthSMP}{0.39\textwidth}
\newlength{\plotwidthDist}
\setlength{\plotwidthDist}{0.6\textwidth}

@@

A~parallel Sudoku solver was used by \citet{par-monad} to compare |Par| Monad
to Multicore Haskell. We ported it to PArrows.

@@

The \ensuremath{\Conid{Par}} monad and Multicore Haskell can be executed on a shared
memory machines only. Although GpH is available on distributed memory
clusters, and newer distributed memory Haskells such as HdpH exist,
current support of distributed memory in PArrows is limited to
Eden. We used the MPI backend of Eden in a distributed memory
setting. However, for shared memory Eden features a ``CP'' backend
that merely copies the memory blocks between distributed heaps. In
this mode Eden still operates in the ``nothing shared'' setting, but
is adapted better to multicore machines. We label this version of Eden
in the plots as ``Eden~CP''.

@@

The PArrows version of \rmtest on a single node of the Glasgow grid
showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
of 64-task PArrows/Eden at 16 real cores version was 13.65, the
efficiency was 85.3\%.  However, if we increase the number of
requested cores to be 32---\ie if we use hyperthreading on 16 real
cores---the speedup does not increase that well. It is merely 15.99
for 32 tasks with PArrows/Eden. It is worse for other backends.  As
for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
hyperthreaded cores and only 13.55 with PArrows/Multicore
Haskell. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
version used here was Eden~CP, the ``share nothing'' SMP build.

@@

\newcommand{\performanceSkelRMSM}[2]{
\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}{17}{\plotwidthSMP}
}

\newcommand{\speedupSkelRMSM}[2]{
\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}{17}{\plotwidthSMP}
}

@@

\olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
  Why is it good?}
The multicore version of our parallel \rmtest benchmark is depicted in
Figure~\ref{fig:bench-rm-sm}. We executed the test with 32 and 64
tasks. The plot shows the PArrows-enabled versions with corresponding backends.
The performance of PArrows/Eden~CP in shared memory is slightly better than
for SMP variants such as PArrows/Multicore Haskell and PArrows/|Par|
Monad but most of the time the performance is still comparable with the Multicore backend performing slightly worse than the other two in terms of speedup. In particular, the speedups for 16 cores fall behind quite noticeably with 11.69 for 32 tasks as Eden and the |Par| Monad both score 13.05 for 32 tasks. While the Multicore backend scores a bit better in the 64 tasks benchmark, its speedup of 12.70 is still behind the 13.65 and 13.33 of Eden and |Par| Monad, respectively.

@@

@@

@@

@@

@@

As the last Benchmark in this paper we present the \sudokutest in \ref{fig:sudokuSMBenchmark} running in a shared memory setting. Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark}-\ref{fig:sudokuSMBenchmark16000}. However, we notice that the Multicore backend seems to choke on a bigger input \ref{fig:sudokuSMBenchmark16000}. This is due to the benchmark only using |parMap| instead of a chunking variant (however we did not change that for simplicity's sake) and is reflected by debug output which shows, that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool. Another remarkable finding is that the Eden backend seems to lack behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.

\newcommand{\performanceSudokuSM}[1]{
\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{32}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
%\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
}{33}{\plotwidthSMP}
}

\newcommand{\speedupSudokuSM}[1]{
\speedupplot{ \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{32}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
%\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
}{33}{\plotwidthSMP}
}

@@

\begin{figure}
\centering
%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
%
{\speedupSudokuSM{16000}\label{subfig:speedupSudokuSM16000}}%
%\hfill%
{\performanceSudokuSM{16000}\label{subfig:performanceSudokuSM16000}}
\caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
  grid; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
  closer to linear speedup the better. The Multicore version shows signs of choking with too many sparks being created.}
\label{fig:sudokuSMBenchmark16000}
\end{figure}

>>>>

\newcommand{\rmtest}{Rabin--Miller test\xspace}
\newcommand{\sudokutest}{Sudoku\xspace}
\newcommand{\jacobitest}{Jacobi sum test\xspace}
\newcommand{\torustest}{Gentleman\xspace}
\newlength{\plotwidthSMP}
\setlength{\plotwidthSMP}{0.39\textwidth}
\newlength{\plotwidthDist}
\setlength{\plotwidthDist}{0.6\textwidth}

@@

A~parallel Sudoku solver was used by \citet{par-monad} to compare |Par| Monad
to GpH Haskell. We ported it to PArrows.

@@

The \ensuremath{\Conid{Par}} monad and GpH Haskell can be executed on a shared
memory machines only. Although GpH is available on distributed memory
clusters, and newer distributed memory Haskells such as HdpH exist,
current support of distributed memory in PArrows is limited to
Eden. We used the MPI backend of Eden in a distributed memory
setting. However, for shared memory Eden features a ``CP'' backend
that merely copies the memory blocks between distributed heaps. In
this mode Eden still operates in the ``nothing shared'' setting, but
is adapted better to multicore machines. We label this version of Eden
in the plots as ``Eden~CP''.

@@

The PArrows version of \rmtest on a single node of the Glasgow grid
showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
of 64-task PArrows/Eden at 16 real cores version was 13.65, the
efficiency was 85.3\%.  However, if we increase the number of
requested cores to be 32---\ie if we use hyperthreading on 16 real
cores---the speedup does not increase that well. It is merely 15.99
for 32 tasks with PArrows/Eden. It is worse for other backends.  As
for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
hyperthreaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
version used here was Eden~CP, the ``share nothing'' SMP build.

@@

\newcommand{\performanceSkelRMSM}[2]{
\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, GpH, |Par| Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}{17}{\plotwidthSMP}
}

\newcommand{\speedupSkelRMSM}[2]{
\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, GpH, |Par| Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}{17}{\plotwidthSMP}
}

@@

\olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
  Why is it good?}
The multicore version of our parallel \rmtest benchmark is depicted in
Figure~\ref{fig:bench-rm-sm}. We executed the test with 32 and 64
tasks. The plot shows the PArrows-enabled versions with corresponding backends.
The performance of PArrows/Eden~CP in shared memory is slightly better than
for SMP variants such as PArrows/GpH and PArrows/|Par|
Monad but most of the time the performance is still comparable with the GpH backend performing slightly worse than the other two in terms of speedup. In particular, the speedups for 16 cores fall behind quite noticeably with 11.69 for 32 tasks as Eden and the |Par| Monad both score 13.05 for 32 tasks. While the GpH backend scores a bit better in the 64 tasks benchmark, its speedup of 12.70 is still behind the 13.65 and 13.33 of Eden and |Par| Monad, respectively.

@@

\renewcommand{\speedupJacobiDist}[4]{
	\textcolor{red}{benchmark missing. in the works}
}

@@

\renewcommand{\performanceJacobiDistDiff}[5]{
	\textcolor{red}{benchmark missing. in the works}
}

@@

\renewcommand{\speedupTorusDist}[3]{
	\textcolor{red}{benchmark missing. in the works}
}

@@

\renewcommand{\performanceTorusDistDiff}[4]{
		\textcolor{red}{benchmark missing. in the works}
}

@@

As the last Benchmark in this paper we present the \sudokutest in \ref{fig:sudokuSMBenchmark} running in a shared memory setting. Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark}-\ref{fig:sudokuSMBenchmark16000}. However, we notice that the GpH backend seems to choke on a bigger input \ref{fig:sudokuSMBenchmark16000}. This is due to the benchmark only using |parMap| instead of a chunking variant (however we did not change that for simplicity's sake) and is reflected by debug output which shows, that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool. Another remarkable finding is that the Eden backend seems to lack behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.

\newcommand{\performanceSudokuSM}[1]{
\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
%\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
}{17}{\plotwidthSMP}
}

\newcommand{\speedupSudokuSM}[1]{
\speedupplot{ Parallel speedup of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
%\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
}{17}{\plotwidthSMP}
}

@@

\begin{figure}
\centering
%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
%
{\speedupSudokuSM{16000}\label{subfig:speedupSudokuSM16000}}%
%\hfill%
{\performanceSudokuSM{16000}\label{subfig:performanceSudokuSM16000}}
\caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
  grid; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
  closer to linear speedup the better. The GpH version shows signs of choking with too many sparks being created.}
\label{fig:sudokuSMBenchmark16000}
\end{figure}

### 2017-09-26 paper/content/conclusion.tex
minor
replaced "Multicore Haskell" with "GpH"

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are the first ones to represent
\emph{parallel} computation with Arrows.
% \olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done
%
Arrows turn out to be a useful tool for composing in parallel
programs. We do not have to introduce new monadic types that wrap the
computation. Instead, we use Arrows in the same manner one uses sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| typeclass for Multicore Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort. It is quite straightforward to add further backends. 
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
%
The benefit is, however, the greatly increased portability of parallel
programs.

\mbcomment{mention ArrowLoop in Torus and Ring chapters}
%\olcomment{Parrows + accelerate = love?} \olcomment{Metion port to Frege.}

>>>>

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are first to represent
\emph{parallel} computation with Arrows.
% \olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done
%
Arrows turn out to be a useful tool for composing in parallel
programs. We do not have to introduce new monadic types that wrap the
computation. Instead, we use Arrows in the same manner one uses sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| typeclass for GpH Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort. It is quite straightforward to add further backends. 
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
%
The benefit is, however, the greatly increased portability of parallel programs.

%\mbcomment{mention ArrowLoop in Torus and Ring chapters}
%\olcomment{Parrows + accelerate = love?} \olcomment{Metion port to Frege.}

### 2017-09-26 paper/content/futures.tex
replaced "Multicore Haskell" with "GpH"

We should allow the nodes to communicate directly with each other. Eden already ships with "remote data" that enable this \cite{AlGo03a,Dieterle2010}.
But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
\begin{figure}[h]
\begin{code}
class Future fut a | a -> fut where
    put :: (Arrow arr) => arr a (fut a)
    get :: (Arrow arr) => arr (fut a) a
\end{code}
\caption{Definition of the |Future| typeclass.}
\label{fig:future}
\end{figure}
Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows. %This is due to the same reason we had to introduce a wrapper for |Strategy a| in the Multicore Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
Technical details are in Appendix, in Section~\ref{app:omitted}.

For our |Par| Monad and Multicore Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}) which are just simple wrappers around the actual data, as in a shared memory setting we do not require Eden's sophisticated communication channels.
\begin{figure}[tb]
\begin{code}
data BasicFuture a = BF a

@@

instance (NFData a) => Future BasicFuture a where
    put = arr BF
    get = arr (\(BF a) -> a)
\end{code}
\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and Multicore Haskell backends.}
\label{fig:BasicFuture}
\end{figure} % $

>>>>

We should allow the nodes to communicate directly with each other. Eden already ships with "remote data" that enable this \cite{AlGo03a,Dieterle2010}.
But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
\begin{figure}[h]
\begin{code}
class Future fut a | a -> fut where
    put :: (Arrow arr) => arr a (fut a)
    get :: (Arrow arr) => arr (fut a) a
\end{code}
\caption{Definition of the |Future| typeclass.}
\label{fig:future}
\end{figure}
Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows. %This is due to the same reason we had to introduce a wrapper for |Strategy a| in the GpH Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
Technical details are in Appendix, in Section~\ref{app:omitted}.

For our |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}) which are just simple wrappers around the actual data, as in a shared memory setting we do not require Eden's sophisticated communication channels.
\begin{figure}[tb]
\begin{code}
data BasicFuture a = BF a

@@

instance (NFData a) => Future BasicFuture a where
    put = arr BF
    get = arr (\(BF a) -> a)
\end{code}
\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and GpH Haskell backends.}
\label{fig:BasicFuture}
\end{figure} % $

### 2017-09-26 paper/content/motivation.tex
replaced "Multicore Haskell" with "GpH"

One particular reason for this is the ease of introducing new sophisticated language
concepts. Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms including the expression of parallelism. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
the Multicore Haskell (a SMP implementation of GpH language), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. 
%In this paper we introduce a notion of parallel
%computations using Arrows.
A key novelty here is to use Arrows to represent parallel computations. They seem a natural fit as they are a generalization of the function |->| and serve as general interface to computations.

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and implementations for three parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface and varying implementations in
existing parallel Haskells: Multicore Haskell, |Par| Monad, and
Eden. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
the parallel Haskell backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

>>>>

One particular reason for this is the ease of introducing new sophisticated language
concepts. Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms including the expression of parallelism. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
the GpH Haskell (a SMP implementation of GpH language), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. 
%In this paper we introduce a notion of parallel
%computations using Arrows.
A key novelty here is to use Arrows to represent parallel computations. They seem a natural fit as they are a generalization of the function |->| and serve as general interface to computations.

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and implementations for three parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface and varying implementations in
existing parallel Haskells: GpH Haskell, |Par| Monad, and
Eden. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
the parallel Haskell backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

### 2017-09-26 paper/content/parallelHaskells.tex
minor
replaced "Multicore Haskell" with "GpH"

\subsection{Short introduction to parallel Haskells}
\label{sec:parEvalNIntro}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel or |parEvalN :: [a -> b] -> [a] -> [b]|, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of |parEvalN|.}
	\label{fig:parEvalN}
\end{figure}

@@

There exist several ways to implement general parallelism in Haskell. Among the most important are probably GpH  (based on |par| \& |pseq| "hints") \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \cite{eden,Loogen2012} (a parallel Haskell for distributed memory), HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.

@@

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.

@@

\begin{figure}[t]
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
	\caption{Dataflow of the Multicore Haskell |parEvalN| version.}
	\label{fig:parEvalNMulticoreImg}
\end{figure} %$ %% formatting

\subsubsection{Par Monad}
The |Par| Monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs. Let:

@@

The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| Monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get |[b]|.
Fig.~\ref{fig:parEvalNParMonadImg} shows a graphical representation.

@@

The trace visualizations are color-coded. In such a visualization (Fig.~\ref{fig:withoutFutures}), the $x$ axis shows the time, the $y$ axis enumerates the machines and processes. The visualization shows a running process in green, a blocked process is red. If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow. The typical reason for thus is GC. An inactive machine, where no processes are started yet, or all are already terminated, shows as a blue bar. A~comminication from one process to another is represented with a black arrow. A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.

>>>>

\subsection{Short introduction to parallel Haskells}
\label{sec:parEvalNIntro}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel or |parEvalN :: [a -> b] -> [a] -> [b]|, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of |parEvalN|. A list of inputs is transformed by different functions in parallel.}
	\label{fig:parEvalN}
\end{figure}

@@

There exist several parallel Haskells already. Among the most important are probably GpH  (based on |par| \& |pseq| "hints") \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \cite{eden,Loogen2012} (a parallel Haskell for distributed memory), HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.

@@

\subsubsection{Glasgow parallel Haskell - GpH}
GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{An implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.

@@

\begin{figure}[t]
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
	\caption{Dataflow of the GpH |parEvalN| version.}
	\label{fig:parEvalNMulticoreImg}
\end{figure} %$ %% formatting

\subsubsection{Par Monad}
The |Par| Monad\footnote{The |Par| monad can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs. Let:

@@

The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with GpH. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| Monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get |[b]|.
Fig.~\ref{fig:parEvalNParMonadImg} shows a graphical representation.

@@

The trace visualizations are color-coded. In such a visualization (Fig.~\ref{fig:withoutFutures}), the $x$ axis shows the time, the $y$ axis enumerates the machines and processes. The visualization shows a running process in green, a blocked process is red. If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow. The typical reason for this is GC. An inactive machine, where no processes are started yet, or all are already terminated, shows as a blue bar. A~comminication from one process to another is represented with a black arrow. A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.
